Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-05-08 Manuscript/PLoSONEresponse/response.tex

We thank the reviewer for pointing out this error. We fixed the shape of the distribution in Figure 3-B.

%-------------

\textcolor{blue}{20-``Combined effect of missing data parameters" section ANOVA or multiple regression are the appropriate analyses to tease apart multiple interacting factors. Here the authors opt to describe results without statistical tests, which is disappointing.}

This comment links back to the reviewer's slight misunderstanding of the statistics we applied in this paper. The purpose of ANOVA or multiple regression is to compare differences in means across groups and test whether they are different from what one would expect according to a null model. In contrast, in our simulation approach, we essentially have access to a representative sample of the entire population distribution of tree topology comparisons. Therefore, we can simply ask whether these distributions are different across XXX [AJ - treatment groups or whatever] by comparing them directly. One standard such approach to comparing distributions directly is to use the BHAttacadsdfasdf coefficient. This number has direct intepretation in a probablistic sense and is simply the probability that the two distributions are the same. If the two distirubions overlap completely then they have BHAVYY coefficient and probability of being the same of 1. In the other extreme, if the two distributions are completely disjointed, then they have a BHATTYV coefficient and hence probabilty of being the same of 0. Thus, the BHYVYVY is a probability, comparable to, but not the same as, a p-value that would from frequentist approaches such as t-tests, anova or glms. Furthermore, as a direct comparison of two known distributions, it is a very simple and test that comes with far fewer assumptions than are made in general linear models and their analogues. For example, the direct comparison approach using BYGYVHVHV is capable of comparing a multi-modal distributions whereas the assumption of normality required in a t-test or glm would inviolate the robustness of the resultant p-value. There is an additional complitation to using t-tests or glms in simulation type analyeses arising from having access to very large smaple sizes that leads to p-values shrinking to zero and resulting in comparisons being almost certain to yield "statistically significant" differences in the mean. We argue that in this situation, it is better to compare the entire distribution using BHCYCY rather than simply the means. [AJ - you could add a simple figure of two normal distributions with 10^4 samples in each that have very similar means. Do a t-test which will be significant, but BHVYVYVYV will be nearly 1.

In our analysis, we made the methodological choice to compare tree similarity distributions that are: not continuous distributions; bounded at 1; not always normal (both the distributions and the residuals) and sometimes multi-modal. Therefore, we choose not to use classical ``statistics" methods such as ANOVA or multiple regression because these methods could simply not be applied to our data set.
%Not comparing means, non normal distributions. And non-parametric version (e.g. Kruskal Wallis) not approriate either.

### 2015-05-08 Manuscript/PLoSONEresponse/response.tex

\textcolor{blue}{Reviewer $\#$ 1: This paper tackles an interesting questions. Despite some problems noted below, the experimental design seems adequate.}

%-------------

\textcolor{blue}{A-Unfortunately the data analysis performed is simply the calculation of a number of summary statistics about the overlap between the estimates returned for different simulation conditions. I suspect that if the authors take some time to do statistics (calculate some P-values, confidence intervals, or take a Bayesian perspective), they'll be able to make a statistical case for most of the points that they are trying to make. But I just don't think that a simulation comparison should be publishable without *some* statistical analysis (beyond the calculation of summaries).}

We feel that the reviewer has slightly misunderstood the statistics we applied in this paper. The purpose of ANOVAs or multiple regression (see suggestion from comment 20 below) is to compare differences in means across groups and test whether they are different from what one would expect according to a null model. In contrast, in our simulation approach, we essentially have access to a representative sample of the entire population distribution of tree topology comparisons. Therefore, we can simply ask whether these distributions are different when our missing data parameters varies by comparing them directly. One standard such approach to comparing distributions directly is to use the Bhattacharyya Coefficient. This number has direct interpretation in a probablistic sense and is simply the probability that the two distributions are the same. If the two distributions overlap completely then they have Bhattacharyya Coefficient and probability of being the same of 1. In the other extreme, if the two distributions are completely disjointed, then they have a Bhattacharyya Coefficient and hence probability of being the same of 0 (line 306 and 307 in the previously submitted manuscript). Thus, the Bhattacharyya Coefficient is a probability, comparable to, but not the same as, a p-value that could be calculated using frequentist approaches such as t-tests, ANOVAs or GLMs.

Furthermore, as a direct comparison of two known distributions, using the the Bhattacharyya Coefficient comes with far fewer statistical assumptions than are required in ANOVAs, GLMs and their analogues. For example, the direct comparison approach using Bhattacharyya Coefficient is capable of comparing a multi-modal distributions whereas the assumption of normality required in a ANOVAs or GLMs thus using them would inviolate the robustness of the resultant p-value. There is an additional complication to using ANOVAs or GLMs in simulation type analyses arising from having access to very large sample sizes (e.g. 50000 data points for the Maximum Likelihood bootstraps trees and the Bayesian posterior distribution trees) that leads to p-values shrinking to zero and resulting in comparisons being almost certain to yield "statistically significant" differences in the mean (see the attached figure ; code for reproducibility is available \href{https://github.com/TGuillerme/Total_Evidence_Method-Missing_data/blob/master/Analysis/Comparing_distribution_example.R}{here} ; https://goo.gl/4djNUf). We argue that in this situation, it is better to compare the entire distribution using Bhattacharyya Coefficient rather than simply the means. Additionally, because we are comparing topologies in this paper, we made the methodological choice to use the modal topology (i.e. the most represented topology) instead of the mean topology (i.e. a hypothetical ``middle'' topology).

Finally, we would like to underline that we report our 50\% and 95\% confidence intervals in Figures 4 and 5 when showing the effect of the missing data parameters (calculation detailed lines 295-295) as well as the $1^st$ and $3^rd$ quartiles when comparing the methods in Table 1.

\includegraphics[scale=0.8, keepaspectratio=true]{Comparing_distributions.pdf} \\
\textit{Different results between the Bhattacharyya Coefficient showing that the distributions are really similar (0.999 overlap probability) and a frequentist approach (simple t-test) underlining a significant difference (p-value=0).}

@@

Lines @@@112 to @@@114 in our revision:

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
Each Bayesian tree was estimated using two runs of four chains each for a maximum of 5$\times$$10^7$ generations. For each estimation, we used the ``true" tree's topology as a starting tree (with a starting value for each branch length of one). We also used two priors on the molecular part of the matrix: an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5, and a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)); and one prior on the morphological part of the matrix (exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5). We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took $~$ 140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes). The detailed MrBayes parameters are available in the Supporting Information S1.
\end{minipage}

\textcolor{blue}{C-The use of Mk rather than Mkv makes the study less relevant to real data analyses.}

%Underline the difference between evolutionary models (M\textit{k}, GTR, JC, HKY) and ``models'' that are methods (usually computer algorithms) used to build trees (such as M\textit{kv})

@@

%Again, in our paper, we use the term M\textit{k} to describe the evolutionary model because some characters could be invariant ($\mu$ $>$ 0). However, we do understand that this is rather a semantic point (model $\neq$ algorithm) and we are confident that the description of the model and the software used in the paper will be enough for a careful reader to understand our protocol. We therefore changed the mentions of M\textit{k} to M\textit{kv} in the revision to make less careful readers understand that we were aware of acquisition bias when we chose our phylogenetic inference software.

%-------------

\textcolor{blue}{D-Many of the equations are incorrect.} %only two in the main body.

We thank the reviewer to underline the two errors in the equations in the main body at lines 258 and 275, see our detailed response to the comments number 15 and 17 below. Also we recognise that the equations in our ``Supporting Information S2: Tree Comparisons'' are confusing and are actually unnecessary for the understanding of this paper. We therefore removed the details about the Robinson-Foulds and the Triplets distances from the ``Supporting Information S2: Tree Comparisons'' section.

@@

See our detailed response to comment A above.

%-------------

\textcolor{blue}{21-line 398: ``mMthods" $->$ ``Methods"}

We thanks the reviewer for pointing out this typo and fixed it.

\textcolor{blue}{22-line 420: contra the statement here, a consensus collapsing to a polytomy will affect the RF distance (could go up or down, but it will change).}

@@

\textcolor{blue}{23-more confusion about the min RF here. Also note that if you are counting the entire leaf set as a clade (e.g the star tree having N=1), then the number of clades in rooted binary tree is n-1 (not n-2). Equation 3 has an extra ``-2" As stated it implies that the max RF distance for n=3 is 0, when the correct answer is 2.}

%check

%-------------

\textcolor{blue}{24-The definition of the scaled RF implies that it is never positive (sinc the numerator cannot be positive and the denominator is positive).}

@@

%not in this paper

%-------------

\textcolor{blue}{25-The decision to use Yule trees as the base line for the mean distance is not explained, and seems odd given that the true tree are not Yule trees. I'm not sure that it makes much difference, but it is odd.}

@@

We cited the reference (Bogdanowicz 2012) of the method.
%explain in a bit more detail

%-------------

\textcolor{blue}{26-The NTS can only -infinity for distances that have the property that the mean distance between random Yule trees is 0. This is not true of either distance metric here. So the real range of values is (mean-max)/mean for both the RF and triples distance.}

@@

As clarified in our revision (lines @@@), the NTS is a generalisation of the topological accuracy method (Price \textit{et al.} 2010 PLoS ONE) and is therefore not specialised for the RF or the Triples distance.

%-------------

\textcolor{blue}{27-Note that JavaScript is a language, so it a bit confusing to refer to a Java program as a ``Java script."}

@@

\textcolor{blue}{28-Equation (6) is not correct. You shouldn't set the summation sign as the left hand side of an equation because the notation has a standard definition. More importantly, the $\#$ of triples is the number of ways of drawing 3 (unordered) leaves from the leaf set. So it is n choose 3 not n choose 4.}

%check

%-------------

\textcolor{blue}{29-Equation (7) is true for a three-leaf tree with a polytomy occurring with probability = 1/4 and the other 3 trees being equiprobable. I don't think that is is true for any other weighting of trees for other numbers of taxa. It might be, but the authors must show this.}

@@

%check

%-------------

\textcolor{blue}{30-I don't think that equations 8, 9, or 10 are correct for any case other than the 3 taxon case (though I do note that the authors are using n choose 3 here).}

### 2015-05-10 Manuscript/PLoSONEresponse/TEM_PLoS_revision.tex
fixed revisions (comment 22)

The effect of the missing data in the fossil record ($M_{F}$) is less than the effect of the $M_{L}$ parameter on clade conservation (Normalised Robinson-Foulds metric) but greater on the displacement of wildcard taxa (Normalised Triplets metric; Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). This is related to the fact that the Bayesian consensus tree is built using a majority consensus rule. When the fossil taxa have less data (e.g. $M_{F}$ = 75\%) they will tend to branch with any taxon in the clade that shares most characters with the fossils. Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees in the Bayesian posterior distribution) and the fossil taxa will form a polytomy at the base of the clade. This does not decrease the Normalised Robinson-Foulds metric, because clades are conserved, but decreases the Normalised Triplets metric because the fossils act as wildcard taxa. Conversely, the same scenario in a Maximum Likelihood framework will lead to a dichotomous branching of the fossils but with low bootstrap support ($<$ 50). In other words, the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree, where the fossil will be placed with lower confidence at a higher taxonomic level. We argue that using the Bayesian consensus tree topology is preferable because it is more conservative (e.g. \cite{pattinsonphylogeny2014}).

>>>>

The effect of the missing data in the fossil record ($M_{F}$) is less than the effect of the $M_{L}$ parameter on clade conservation (Normalised Robinson-Foulds metric) but greater on the displacement of wildcard taxa (Normalised Triplets metric; Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). This is related to the fact that the Bayesian consensus tree is built using a majority consensus rule. When the fossil taxa have less data (e.g. $M_{F}$ = 75\%) they will tend to branch with any taxon in the clade that shares most characters with the fossils. Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees in the Bayesian posterior distribution) and the fossil taxa will form a polytomy at the base of the clade. 
In this case, the Normalised Robinson-Foulds metric will decreases when the fossil are present at a lower taxonomic level (i.e. separated with more nodes from the root) but affects the clade conservation less at higher taxonomic level (i.e. separated with less nodes from the root). On the opposite though, because the fossil in a high level clade has many chances to branch on different nodes within that clade, it will be more likely to act as a wildcard taxa and decrease the Normalised Triplets metric. Therefore, the $M_{F}$ parameter is likely to affect the Normalised Robinson-Foulds metric less than the Normalised Triplets metric for the Bayesian consensus trees.
%This does not decrease the Normalised Robinson-Foulds metric, because clades are conserved, but decreases the Normalised Triplets metric because the fossils act as wildcard taxa.
Conversely, the same scenario in a Maximum Likelihood framework will lead to a dichotomous branching of the fossils but with low bootstrap support ($<$ 50). In other words, the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree, where the fossil will be placed with lower confidence at a higher taxonomic level. We argue that using the Bayesian consensus tree topology is preferable because it is more conservative (e.g. \cite{pattinsonphylogeny2014}).
